Modeling workflow: NLA N2O survey data

### 2023-10-16 scripts/NLA_N2O_models.Rmd
Updated files for modeling workflow documentation

---
title: "Modeling workflow: Lake N2O survey data"
author: "Roy Martin, Jake Beaulieu, Michael McManus"
date: "`r Sys.Date()`"
output:
  github_document:
    number_sections: TRUE
    df_print: "tibble"
    math_method: 
      engine: webtex
      url: https://latex.codecogs.com/svg.image?
    html_preview: TRUE
    keep_html: TRUE
bibliography: RWM_Endnote_Library.bib
link-citations: yes
editor_options:
  chunk_output_type: inline
---

```{r begin, eval=TRUE, include=FALSE}
library(ggpubr)
library(moments)
library(ggplot2)
library(ggExtra)
library(gridExtra)
library(kableExtra)
library(ggrepel)
library(dplyr)
library(tidyverse)
library(tidyr)
library(future)
library(foreach)
library(itertools)
library(bayesplot)
library(tidybayes)
library(brms)

@@

# function for DHARMa residual analysis
check_brms <- function(model,             # brms model
                       integer = FALSE,   # integer response? (TRUE/FALSE)
                       plot = TRUE,       # make plot?
                       resp = NULL,
                       ...                # further arguments for DHARMa::plotResiduals 
) {

  mdata <- brms::standata(model)
  if(!"Y" %in% names(mdata))
    oResp <- mdata[[paste0("Y_", resp)]]
  else
    oResp <- mdata[["Y"]]
  #  stop("Cannot extract the required information from this brms model")

  dharma.obj <- DHARMa::createDHARMa(
    simulatedResponse = t(brms::posterior_predict(model, resp = resp, ndraws = 1000)),
    observedResponse = oResp, 
    fittedPredictedResponse = apply(
      t(brms::posterior_epred(model, resp = resp, ndraws = 1000, re.form = NA)),
      1,
      median),
    integerResponse = integer)

  if (isTRUE(plot)) {
    plot(dharma.obj, ...)
  }

  invisible(dharma.obj)

}
```

# Background and Objectives
This document details the modeling workflow implemented for estimating dissolved and equilibrium N2O gas concentrations and saturation ratios for all freshwater lakes and reservoirs in the lower 48 US states larger than 4 ha. Data for these estimates came from the 2017 Nation Lakes Assessment (NLA) survey, wherein waterbodies were sampled according to a spatially balanced, stratified, and unequal probability design. Stratification was among categories of lake size (surface area in hectares), WSA9 ecoregion, and US state (excluding AK and HI); and larger lakes were intentionally over-sampled relative to smaller lakes. In order for this sample to be useful for making inferences relevant to the population of interest, any estimates derived from it needed to be adjusted for potential biases arising from these design factors. It can be useful to consider the complete data likelihood when imagining how such biases may arise in sampling designs[@Link_Barker_2010; @Zachmann_etal_2022]. For the 2017 NLA survey, the complete data was considered to be all US lakes and reservoirs in the lower 48 states larger than 4 hectares. The survey samples, by comparison, are then considered a subset of the complete data with a known pattern of "missingess" assigned by the survey. For example, some lakes in the target population were missing from the sample not at random, but conditional on the design parameters. Their "missingness" is not random and, therefore, not ignorable when making inferences from the sample to the population of interest [@Gelman_etal_2014, Ch. 8; @Zachmann_etal_2022].

One model-based strategy to account for this selection bias is to include the survey design variables as predictors in a regression model and then adjust the resulting estimates based on a known target population distribution; a post-fitting adjustment referred to as poststratification [@Gelman_etal_2020, Ch. 17]. A particularly popular version of this approach uses multilevel regression models and is aptly referred to as multilevel regression and poststratification, or MRP [@Park_etal_2004; @Gelman_Little_1997; @Gelman_etal_2020, Ch. 17; @Kennedy_Gelman_2021]. Multilevel models are often recommended because they can provide regularized estimates along the design groupings, which can improve out-of-sample inferences [e.g., @Gelman_Little_1997; @Kennedy_Gelman_2021]. Estimates for group levels that may be missing from the sample, but are part of the population of interest, are also straightforward using the multilevel approach [@Gelman_etal_2020 Ch. 17; @McElreath_2020].

With a stratified, unequal probability design, such as the one used for the 2017 NLA, the selection design prescribes a number of random draws for each of a number of pre-specified groupings (i.e., sampling unit types). Within a particular stratification level or grouping, the draws may be considered a simple random sample (SRS) from that group type in the target population. A regression model conditioning the sample observations on the design grouping structure could, therefore, be parameterized to provide group-specific estimates that are unbiased with respect to the target population. To make inferences to the broader target population, however, those regression estimates would need to be re-weighted by the known proportions of units across groups in the target population. If the estimates are not adjusted in this manner, the estimates may be biased, particularly if the response of interest varies importantly across the grouping structure. For example, for the 2017 NLA survey, the proportion of samples prescribed to small lakes was much lower, relative to their proportion in the target population, compared to larger lakes, which were sampled in greater proportion relative to the target population. Small lakes are so numerous in the US that, if this unequal weighting by lake size weren't prescribed, nearly all the samples from the survey would be from small lakes and inferences regarding larger lakes could be highly uncertain for any reasonably achievable number of samples.  On the other hand, if lake size had no real influence on the N2O responses, the stratification across size class would make little difference with regard to inferences. Sampling any sizes of lakes would result in the same marginal estimates and standard errors.

For a recent applied example and helpful tutorial employing these concepts in a model-based approach using MRP, see @Kennedy_Gelman_2021. For an example of a recent application in the context of national surveys of environmental resources, see @Zachmann_etal_2022. In this study, a similar model-based approach was used to make population-level estimates from the 2017 NLA dissolved gas data. The specific workflow, data, models, and code used to make these estimates are documented in the remainder of this document. First, Bayesian multilevel regression models were used to fit the sample data to the design variable structure. Next, because eventual flux estimates needed to be estimated from lake-level measures (i.e., surface area), instead of predicting to a typical postratification table, predictions were made to each individual lake in the population of interest. This meant predicting to 465,897 individual natural and man made US lakes larger than 4 hectares in the lower 48 states. These predictions were assumed relevant to average conditions during the biological index period for each lake in 2017. The specific objective of the modeling effort was to provide population estimates for (1) dissolved and equilibrium N2O concentrations; (2) the N2O saturation ratio (i.e., dissolved N2O/equilibrium N2O); and (3) the proportion of under-saturated water bodies (i.e., saturation ratio < 1). As previously indicated, the gas estimates would also be used to estimate the total flux of N2O gas attributable to the target population of lakes over the index period. Saturation ratio was calculated as the ratio of dissolved to equilibrium N2O. Because dissolved and equilibrium N2O were observed on the same sample units (lake sites), models were fit to their joint distribution and the ratio estimates were assembled as a derived quantity. The response variable was, therefore, multivariate in order to account for any potential statistical dependencies between dissolved and equilibrium N2O due to, for example, common dependencies on geography. Although predictions of the mean marginal probabilities from separate models may have provided comparable estimates, a joint model allowing correlated errors was expected to better capture uncertainty and potentially improve out-of-sample predictions, should the variables be conditionally correlated [@Warton_etal_2015; @Poggiato_etal_2021]. All of the models fit were constructed using the `brms` package [@Burkner_2017] in `R` [@R_Core_Team_2021] as an interface to Stan, a software package for fitting fully Bayesian models via Hamiltonian Monte Carlo [HMC, @Stan_Development_Team_2018_a; @Stan_Development_Team_2018_b; @Stan_Development_Team_2018_c]. More specific details on the model structures is provided throughout the "Model fitting" section below.

@@

Gas samples were analyzed via gas chromotography and concentrations were recorded to the nearest 0.001 nmol/L. As part of the survey design, each gas observation was indexed to an individual lake selected with unequal probability from 5 different lake size categories, $j \in j=1,...,J = 5$, according to surface area (ha), and from within a state, $k \in k=1,...,K = 48$, situated within an aggregated, WSA9 or Omernik ecoregion, $l \in l=1,...,L = 9$. All 9 WSA9 ecoregions were represented in the sample, including Xeric (XER), Western Mountain (WMT),  Northern Plains (NPL), Southern Plains (SPL), Temperate Plains (TPL), Coastal Plains (CPL), Upper Midwest (UMW), Northern Appalachian (NAP), and Southern Appalachian (SAP) regions. As shown below, the data from the initial and revisit samples were separately compiled into data frame objects in $\textbf{R}$, with $n=984$ and $n=95$ rows, respectively, of gas observations indexed to the survey design variables and several potentially relevant covariates.

## Import
The gas data and covariates were previously described and munged at 
https://github.com/USEPA/DissolvedGasNla/blob/master/scripts/dataMunge.html. That dataset was imported below.

@@

A new data frame for modeling was constructed from the original file including only the variables of interest: (1) the N2O gas observations; (2) the survey design variables indexed to those observations; and (3) additional covariates considered potentially useful for improving the fit of the model. The modeling data frame below excluded the second-visit observations, which would later be used for model checking. Some variables from the imported data were renamed for convenience. In addition, the NO3 covariate was rounded according to the documented measurement precision and an alternative version was also created by log-transforming and re-coding the variable as an ordered factor with five levels. The left-most cut point separated observations below the detection limit from the completely observed samples. The remaining cut points were drawn at approximately equal distances in the positive direction along the log scale. Finally, one lake had no N2O gas information and was removed from the data frame.

```{r model_data, echo=TRUE, paged.print=TRUE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/dg.rda")

dg %>%
  filter(!is.na(dissolved.n2o.nmol)) %>% # 1 obs with missing measurement
  nrow() # number of observations before filtering

df_model <- dg %>%
  filter(!is.na(dissolved.n2o.nmol)) %>%
  filter(sitetype == "PROB") %>% # probability samples only
  filter(visit.no == 1) %>%
  mutate(n2o = round(dissolved.n2o.nmol, 2),
         n2o_eq = round(sat.n2o.nmol, 2),
         n2o_sat = n2o.sat.ratio,
         n2o_em = e.n2o.nmol.d,
         n2o_flux = f.n2o.m.d,
         WSA9 = factor(ag.eco9),
         state = factor(state.abb[match(state.nm, state.name)]),
         area_ha = area.ha,
         log_area = log(area_ha),
         chla = chla.result,
         log_chla = log(chla),
         elev = elevation,
         log_elev = log(elev + 1),
         do_surf = o2.surf,
         log_do = log(do_surf),
         bf_max = max.bf,
         sqrt_bf = sqrt(bf_max),
         size_cat = recode(area.cat6, 
                           "(1,4]" = "min_4" ,
                           "(10,20]" = "10_20",
                           "(20,50]" = "20_50",
                           "(4,10]" = "4_10",
                           ">50" = "50_max")) %>%
  mutate(size_cat = factor(size_cat,
                           levels = c("min_4", "4_10", "10_20", "20_50", "50_max"),
                           ordered = TRUE)) %>%
  mutate(no3 = ifelse(nitrate.n.result <= 0.0005, 0.0005, round(nitrate.n.result, 4))) %>%# 1/2 mdl 0.01
  mutate(no3_cat = cut(log(no3), # convert no3 to ordered factor with 5 levels
                       breaks = c(-Inf, -7.5, -5.5, -3.5, -1.5, Inf),
                       labels =seq(1, 5, 1))) %>%
  mutate(no3_cat = factor(no3_cat,
                          levels = seq(1, 5, 1),
                          ordered = TRUE)) %>%
  mutate(date = as.Date(date.col)) %>%
  mutate(jdate = as.numeric(format(date, "%j"))) %>% 
  mutate(lat = map.lat.dd,
         lon = map.lon.dd) %>% # longitude
  mutate(surftemp = surftemp,
         log_surftemp = log(surftemp)) %>% 
  select(WSA9,
         state,
         size_cat,
         site.id,
         lat,
         lon,
         date,
         jdate,
         surftemp,
         log_surftemp,
         area_ha,
         log_area,
         elev,
         log_elev,
         chla,
         log_chla,
         do_surf,
         log_do,
         bf_max,
         sqrt_bf,
         n2o,
         n2o_eq,
         no3,
         no3_cat
         )

save(df_model, file = "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/df_model.rda")

nrow(df_model) # number of obs after filtering

@@

A second data frame including only the second visit observations was constructed below. These data were later used as a "test set" to assess the out-of-sample performance of the model developed on the first-visit or "training set".

```{r test_data, echo=TRUE, paged.print=TRUE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/dg.rda")

# number of observations before filtering probability samples
dg %>%
  filter(!is.na(dissolved.n2o.nmol)) %>% # remove obs with missing response measurements
  nrow()

df_test <- dg %>%
  filter(!is.na(dissolved.n2o.nmol)) %>%
  filter(sitetype == "PROB") %>% # probability samples only
  filter(visit.no == 2) %>%
  mutate(n2o = round(dissolved.n2o.nmol, 2),
         n2o_eq = round(sat.n2o.nmol, 2),
         n2o_sat = n2o.sat.ratio,
         n2o_em = e.n2o.nmol.d,
         n2o_flux = f.n2o.m.d,
         WSA9 = factor(ag.eco9),
         state = factor(state.abb[match(state.nm, state.name)]),
         area_ha = area.ha,
         log_area = log(area_ha),
         chla = chla.result,
         log_chla = log(chla),
         elev = elevation,
         log_elev = log(elev + 1),
         do_surf = o2.surf,
         log_do = log(do_surf),
         bf_max = max.bf,
         sqrt_bf = sqrt(bf_max),
         size_cat = recode(area.cat6, 
                           "(1,4]" = "min_4" ,
                           "(10,20]" = "10_20",
                           "(20,50]" = "20_50",
                           "(4,10]" = "4_10",
                           ">50" = "50_max")) %>%
  mutate(size_cat = factor(size_cat,
                           levels = c("min_4", "4_10", "10_20", "20_50", "50_max"),
                           ordered = TRUE)) %>%
  mutate(no3 = ifelse(nitrate.n.result <= 0.0005, 0.0005, round(nitrate.n.result, 4))) %>%# 1/2 mdl 0.01
  mutate(no3_cat = cut(log(no3), # convert no3 to ordered factor with 5 levels
                       breaks = c(-Inf, -7.5, -5.5, -3.5, -1.5, Inf),
                       labels =seq(1, 5, 1))) %>%
  mutate(no3_cat = factor(no3_cat,
                          levels = seq(1, 5, 1),
                          ordered = TRUE)) %>%
  mutate(date = as.Date(date.col)) %>%
  mutate(jdate = as.numeric(format(date, "%j"))) %>% 
  mutate(lat = map.lat.dd,
         lon = map.lon.dd) %>% # longitude
  mutate(surftemp = surftemp,
         log_surftemp = log(surftemp)) %>% 
  select(WSA9,
         state,
         size_cat,
         site.id,
         lat,
         lon,
         date,
         jdate,
         surftemp,
         log_surftemp,
         area_ha,
         log_area,
         elev,
         log_elev,
         chla,
         log_chla,
         do_surf,
         log_do,
         bf_max,
         sqrt_bf,
         n2o,
         n2o_eq,
         no3,
         no3_cat
         )

save(df_test, file = "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/df_test.rda")

nrow(df_test) # number of obs after filtering for probability samples, first visits, and removing one site missing ecoregion (WSA9) info.

@@

## Target population
Below. the NLA sampling frame was imported and filtered to include all lakes the target population. The resulting target population above included a total of 465,897 waterbodies.

```{r import_sample_frame, echo=TRUE}
df_pop <- read.csv(file = paste0(localPath,
              "/Environmental Protection Agency (EPA)/",
              "ORD NLA17 Dissolved Gas - Documents/",
              "inputData/NLA_Sample_Frame.csv"), header = T)

@@

save(sframe, file = "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/sframe.rda")

sframe %>%
  head(10)
```

Cross tabulations below describe the structure of the target population with respect to the survey design variables. The cross-tabulation shows that each ecoregion does not contain each state. Therefore, in the statistical sense, states were nested in ecoregions.

>>>>

---
title: "Modeling workflow: NLA N2O survey data"
author: "Roy Martin, Jake Beaulieu, Michael McManus"
date: "`r Sys.Date()`"
output:
  github_document:
    number_sections: TRUE
    df_print: "tibble"
    math_method: 
      engine: webtex
      url: https://latex.codecogs.com/svg.image?
    html_preview: TRUE
    keep_html: TRUE
bibliography: RWM_Endnote_Library.bib
link-citations: yes
editor_options:
  chunk_output_type: inline
---

```{r setup, echo = FALSE, warning=FALSE, message=FALSE}
library(ggpubr)
library(moments)
library(ggplot2)
library(ggExtra)
library(gridExtra)
library(kableExtra)
library(ggrepel)
library(dplyr)
library(tidyverse)
library(tidyr)
library(future)
library(foreach)
library(itertools)
library(bayesplot)
library(tidybayes)
library(brms)
library(tictoc)
library(arrow)

@@

# Load model and data files for knitting (if they already exist)
f1 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/dg.rda"
f2 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/df_model.rda"
f3 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/df_test.rda"
f4 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/sframe.rda"
f5 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/pframe.rda"
f6 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/samp_props.rda"

f7 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/n2o_mod1.rda"
f8 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/n2o_mod2.rda"
f9 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/n2o_mod3.rda"
f10 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/n2o_mod4.rda"
f11 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/n2o_mod5.rda"
f12 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/n2o_mod6.rda"
f13 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/all_predictions_b.rda"
f14 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/sframe.rda"
f15 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/pframe.rda"
f16 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/n2o_survey_ests.rda"
f17 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/n2o_survey_ests_size.rda"
f18 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/sat_survey_ests.rda"
f19 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/sat_survey_ests_size.rda"

if(file.exists(f1))
  load(f1)
if(file.exists(f2))
  load(f2)
if(file.exists(f3))
  load(f3)
if(file.exists(f4))
  load(f4)
if(file.exists(f5))
  load(f5)
if(file.exists(f6))
  load(f6)
if(file.exists(f7))
  load(f7)
if(file.exists(f8))
  load(f8)
if(file.exists(f9))
  load(f9)
if(file.exists(f10))
  load(f10)
if(file.exists(f11))
  load(f11)
if(file.exists(f12))
  load(f12)
if(file.exists(f13))
  load(f13)
if(file.exists(f14))
  load(f14)
if(file.exists(f15))
  load(f15)
if(file.exists(f16))
  load(f16)
if(file.exists(f17))
  load(f17)
if(file.exists(f18))
  load(f18)
if(file.exists(f19))
  load(f19)
```

# Background and Objectives
This document describes the modeling workflow and results from estimating dissolved and equilibrium N2O gas concentrations and saturation ratios for all freshwater waterbodies (natural and man-made lakes and reservoirs) in the lower 48 US states larger than 1 ha. Data for these estimates came from the 2017 Nation Lakes Assessment (NLA) survey, wherein a subset of waterbodies were sampled from the broader population of interest (POI) according to a spatially balanced, stratified, and unequal probability design. Stratification was among categories of lake size (surface area in hectares), WSA9 ecoregion, and US state (excluding AK and HI). In order to better capture variability among the wide range of lake sizes in the POI, larger lakes were intentionally over-sampled relative to smaller lakes.

To make inferences back to the POI from this sample, estimates needed to be adjusted in a manner that addressed potential biases related to the relative representation of waterbody characteristics in the sample compared to the POI, specifically with respect to the survey design variables. It can be helpful to define the complete data likelihood (CDL) when considering how to address biases that may arise from sampling designs [@Link_Barker_2010; @Zachmann_etal_2022]. For the 2017 NLA survey, the complete data would be considered all US lakes and reservoirs in the lower 48 states larger than 1 hectare (_i.e._, the POI). The survey samples, by comparison, could then be considered a subset of that complete data with a known pattern of "missingess" assigned by the survey. In this case, waterbodies in the POI were missing from the sample not completely at random, but according to a probability assigned by the design parameters. This type of missingness is considered not ignorable when making inferences from a sample to a POI [@Gelman_etal_2014, Ch. 8; @Zachmann_etal_2022]. However, _conditional_ on the design variables, the sample could be considered a simple random sample (SRS) from the POI [@Gelman_etal_2014, Ch. 8; @Zachmann_etal_2022].

A simple, model-based strategy for adjusting a sample to a POI is to include the survey design variables as predictors in a regression model [@Gelman_etal_2020, Ch. 17]. After fitting the regression, the parameter estimates can be used to make adjustments according to the distribution of the design variables in the POI in a second step referred to as "poststratification" [@Gelman_etal_2020, Ch. 17]. A particularly popular version of this approach uses multilevel regression models and is referred to as multilevel regression with poststratification, or MRP [@Park_etal_2004; @Gelman_Little_1997; @Gelman_etal_2020, Ch. 17; @Kennedy_Gelman_2021]. Multilevel regression models are often recommended because they can provide regularized estimates along the design groupings, which can improve out-of-sample inferences [_e.g._, @Gelman_Little_1997; @Kennedy_Gelman_2021]. Estimates for group levels that may be missing from the sample, but are part of the population of interest, are also straightforward using the multilevel approach [@Gelman_etal_2020 Ch. 17; @McElreath_2020]. For a recent applied example and helpful tutorial employing this approach, see @Kennedy_Gelman_2021. For an example of a recent application in the context of national surveys of environmental resources, see @Zachmann_etal_2022.

In this study, we used a MRP-like approach to provide population-level estimates based on the 2017 NLA dissolved gas survey data. The remainder of this document provides the workflow, data, models, and code used to make these estimates. To begin, a number of Bayesian multilevel regression models were fit to the sample data. These models were fit and checked against both the data used to construct the models as well as hold-out data from future visits. After evaluating a number of different models, a final model was selected for use in inference. Because total N2O flux estimates needed to be estimated based on individual, waterbody-level measures (_e.g._, surface area), predictions from the model were made to each individual lake in the POI (n = 465,897). These predictions were assumed relevant to average conditions during the biological index period for each lake in 2017.

The objective of the modeling effort was to provide estimates for (1) dissolved and equilibrium N2O concentrations; (2) the N2O saturation ratio (i.e., dissolved N2O/equilibrium N2O); and (3) the proportion of under-saturated water bodies (i.e., saturation ratio < 1) for the National-level POI and a number of sub-populations therein (_e.g._, ecoregion, state, size category). The gas estimates were also used to estimate the total flux of N2O gas attributable to the target populations. Saturation ratio was calculated as the ratio of dissolved to equilibrium N2O. Because dissolved and equilibrium N2O were observed on the same sample units (sites in lakes), the regression models were fit to their joint distribution and the ratio estimates were assembled as a derived quantity calculated from the joint posterior predictive distribution of those variables. The response variable was, therefore, modeled as multivariate in order to account for any potential statistical dependencies between dissolved and equilibrium N2O due to, for example, common dependencies on geography. Although predictions of the mean marginal probabilities from separate models may have provided comparable estimates, a joint model allowing correlated errors was expected to better capture uncertainty and potentially improve out-of-sample predictions, should the variables be conditionally correlated [@Warton_etal_2015; @Poggiato_etal_2021]. All of the models fit were constructed using the `brms` package [@Burkner_2017] in `R` [@R_Core_Team_2021] as an interface to Stan, a software package for fitting fully Bayesian models via Hamiltonian Monte Carlo [HMC, @Stan_Development_Team_2018_a; @Stan_Development_Team_2018_b; @Stan_Development_Team_2018_c]. More specific details on the model structures is provided throughout the "Model fitting" section below.

@@

Water samples were analyzed via gas chromotography and concentrations were recorded to the nearest 0.001 nmol/L. As part of the survey design, each gas observation was indexed to an individual lake selected with unequal probability from 5 different lake size categories, $j \in j=1,...,J = 5$, according to surface area (ha), and from within a state, $k \in k=1,...,K = 48$, situated within an aggregated, WSA9 or Omernik ecoregion, $l \in l=1,...,L = 9$. All WSA9 ecoregions were represented, including the Xeric (XER), Western Mountain (WMT),  Northern Plains (NPL), Southern Plains (SPL), Temperate Plains (TPL), Coastal Plains (CPL), Upper Midwest (UMW), Northern Appalachian (NAP), and Southern Appalachian (SAP) regions. Gas data from the first and second (follow-up) visits were separated, with $n=984$ _vs._ $n=95$ observations each. In addition to the design variables, a number of potentially useful covariates were also indexed to the gas observations, including measures of NO3, surface water temperature, elevation, chlorophyll a content, dissolved oxygen content, and waterbody size (hectares). These gas data and covariates were previously described and munged at: https://github.com/USEPA/DissolvedGasNla/blob/master/scripts/dataMunge.html.

## Import
That originally munged gas dataset was imported below.

@@

A new data frame for modeling was constructed from the original file, including only the variables of interest: (1) the N2O gas observations; (2) the survey design variables indexed to those observations; and (3) additional covariates considered potentially useful for improving the fit and/or practical interpretation of the model. The modeling data frame below excluded the second-visit observations, which would later be used for model checking. Some variables from the imported data were renamed for convenience. In addition, the NO3 covariate was rounded according to the documented measurement precision and an alternative version was created by log-transforming and re-coding the variable as an ordered factor with five levels. The left-most cut point separated observations below the detection limit from the completely observed samples. The remaining cut points were drawn at approximately equal distances in the positive direction along the log scale. Finally, one lake had no N2O gas information and was removed from the data frame.

```{r all_data_n, echo=TRUE, paged.print=TRUE}
dg %>%
  filter(!is.na(dissolved.n2o.nmol)) %>% # 1 obs with missing N2O measurement
  nrow() # number of observations before filtering
```

```{r model_data, eval=FALSE, include=TRUE}
df_model <- dg %>%
  filter(!is.na(dissolved.n2o.nmol)) %>%
  filter(sitetype == "PROB") %>% # probability samples only
  filter(visit.no == 1) %>% # first-visit sites only
  mutate(n2o = round(dissolved.n2o.nmol, 2),
         n2o_eq = round(sat.n2o.nmol, 2),
         n2o_sat = n2o.sat.ratio,
         n2o_em = e.n2o.nmol.d,
         n2o_flux = f.n2o.m.d,
         WSA9 = factor(ag.eco9),
         state = factor(state.abb[match(state.nm, state.name)]),
         area_ha = area.ha,
         log_area = log(area_ha),
         chla = chla.result,
         log_chla = log(chla),
         elev = elevation,
         log_elev = log(elev + 1),
         do_surf = o2.surf,
         log_do = log(do_surf),
         bf_max = max.bf,
         sqrt_bf = sqrt(bf_max),
         size_cat = recode(area.cat6, # simpler naming conventions for the size categories
                           "(1,4]" = "min_4" ,
                           "(10,20]" = "10_20",
                           "(20,50]" = "20_50",
                           "(4,10]" = "4_10",
                           ">50" = "50_max")) %>%
  mutate(size_cat = factor(size_cat,
                           levels = c("min_4", "4_10", "10_20", "20_50", "50_max"),
                           ordered = TRUE)) %>%
  mutate(no3 = ifelse(nitrate.n.result <= 0.0005, 0.0005, round(nitrate.n.result, 4))) %>%# 1/2 mdl 0.01
  mutate(no3_cat = cut(log(no3), # convert no3 to ordered factor with 5 levels
                       breaks = c(-Inf, -7.5, -5.5, -3.5, -1.5, Inf),
                       labels =seq(1, 5, 1))) %>%
  mutate(no3_cat = factor(no3_cat,
                          levels = seq(1, 5, 1),
                          ordered = TRUE)) %>%
  mutate(date = as.Date(date.col)) %>%
  mutate(jdate = as.numeric(format(date, "%j"))) %>% 
  mutate(lat = map.lat.dd,
         lon = map.lon.dd) %>% # longitude
  mutate(surftemp = surftemp,
         log_surftemp = log(surftemp)) %>% 
  select(WSA9,
         state,
         size_cat,
         site.id,
         lat,
         lon,
         date,
         jdate,
         surftemp,
         log_surftemp,
         area_ha,
         log_area,
         elev,
         log_elev,
         chla,
         log_chla,
         do_surf,
         log_do,
         bf_max,
         sqrt_bf,
         n2o,
         n2o_eq,
         no3,
         no3_cat
         )

save(df_model, file = "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/df_model.rda") 
```

A preview showing the first 10 rows of the modeling data:

```{r model_data_print, echo=TRUE, paged.print=TRUE}
nrow(df_model) # number of obs after filtering

@@

A second data frame, including only the second visit observations was constructed below. These data were later used as a "test set" to assess the out-of-sample performance of the model developed on the first-visit or "training set".

```{r all_data_n_2, echo=TRUE, paged.print=TRUE}
# number of observations before filtering probability samples
dg %>%
  filter(!is.na(dissolved.n2o.nmol)) %>% # remove obs with missing response measurements
  nrow()
```

```{r test_data, eval=FALSE, include=TRUE}
df_test <- dg %>%
  filter(!is.na(dissolved.n2o.nmol)) %>%
  filter(sitetype == "PROB") %>% # probability samples only
  filter(visit.no == 2) %>% # second-visit sites only
  mutate(n2o = round(dissolved.n2o.nmol, 2),
         n2o_eq = round(sat.n2o.nmol, 2),
         n2o_sat = n2o.sat.ratio,
         n2o_em = e.n2o.nmol.d,
         n2o_flux = f.n2o.m.d,
         WSA9 = factor(ag.eco9),
         state = factor(state.abb[match(state.nm, state.name)]),
         area_ha = area.ha,
         log_area = log(area_ha),
         chla = chla.result,
         log_chla = log(chla),
         elev = elevation,
         log_elev = log(elev + 1),
         do_surf = o2.surf,
         log_do = log(do_surf),
         bf_max = max.bf,
         sqrt_bf = sqrt(bf_max),
         size_cat = recode(area.cat6, # simpler naming conventions for the size categories
                           "(1,4]" = "min_4" ,
                           "(10,20]" = "10_20",
                           "(20,50]" = "20_50",
                           "(4,10]" = "4_10",
                           ">50" = "50_max")) %>%
  mutate(size_cat = factor(size_cat,
                           levels = c("min_4", "4_10", "10_20", "20_50", "50_max"),
                           ordered = TRUE)) %>%
  mutate(no3 = ifelse(nitrate.n.result <= 0.0005, 0.0005, round(nitrate.n.result, 4))) %>%# 1/2 mdl 0.01
  mutate(no3_cat = cut(log(no3), # convert no3 to ordered factor with 5 levels
                       breaks = c(-Inf, -7.5, -5.5, -3.5, -1.5, Inf),
                       labels =seq(1, 5, 1))) %>%
  mutate(no3_cat = factor(no3_cat,
                          levels = seq(1, 5, 1),
                          ordered = TRUE)) %>%
  mutate(date = as.Date(date.col)) %>%
  mutate(jdate = as.numeric(format(date, "%j"))) %>% 
  mutate(lat = map.lat.dd,
         lon = map.lon.dd) %>% # longitude
  mutate(surftemp = surftemp,
         log_surftemp = log(surftemp)) %>% 
  select(WSA9,
         state,
         size_cat,
         site.id,
         lat,
         lon,
         date,
         jdate,
         surftemp,
         log_surftemp,
         area_ha,
         log_area,
         elev,
         log_elev,
         chla,
         log_chla,
         do_surf,
         log_do,
         bf_max,
         sqrt_bf,
         n2o,
         n2o_eq,
         no3,
         no3_cat
         )

save(df_test, file = "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/df_test.rda") 
```

A preview of the first 10 rows of the test data:

```{r test_data_print, echo=TRUE, paged.print=TRUE}
nrow(df_test) # number of obs after filtering for probability samples, first visits, and removing one site missing ecoregion (WSA9) info.

@@

## Target population

Below. the original NLA sampling frame was imported and filtered to include only lakes in the POI. The resulting target population included a total of 465,897 waterbodies.

```{r import_sample_frame, eval=FALSE, include=TRUE}
df_pop <- read.csv(file = paste0(localPath,
              "/Environmental Protection Agency (EPA)/",
              "ORD NLA17 Dissolved Gas - Documents/",
              "inputData/NLA_Sample_Frame.csv"), header = T)

@@

save(sframe, file = "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/sframe.rda") 
```

Below is a sample of the first ten rows describing the POI:

```{r sample_frame_print, echo=TRUE, paged.print=TRUE}
sframe %>%
  head(10)
```

Cross tabulations below show the structure of the target population with respect to the survey design variables. The cross-tabulation shows that each ecoregion does not contain each state. Therefore, states were nested within ecoregions.
